---
title: "Measuring Social Inequality"
subtitle: "Advanced Social Epidemiology PhD Course"  
author: "Sam Harper"
institute: " <br> </br>"
date: "SER Pre-Conference Workshop <br> 2021-10-11 to 2021-10-15 </br>"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, style.css]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(here)
library(tidyverse)
library(DiagrammeR)
library(xaringan)
library(leaflet)
library(ggplot2)
library(emojifont)
library(countdown)
xfun::pkg_load2(c('tikzDevice', 'magick', 'pdftools'))
```

```{r, include=FALSE}
pdf2png = function(path) {
  # only do the conversion for non-LaTeX output
  if (knitr::is_latex_output()) return(path)
  path2 = xfun::with_ext(path, "png")
  img = magick::image_read_pdf(path)
  magick::image_write(img, path2, format = "png")
  path2
}
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#000000", header_color = "#737373", text_font_size = "24px",  text_font_family = "'Lucida Sans'", header_font_google = google_font("Source Sans Pro"), header_font_weight="lighter", title_slide_background_color =  "#ffffff", title_slide_text_color = "#000000", link_color = "#0000ee", footnote_font_size = "0.5em")
```

class: center, middle, inverse
# .orange[**Part 2: Measuring the "Inequality" in Social Inequality**]

---
class: center, top, inverse
# .orange[**2. Measuring Inequality**]

.left[
## .orange[**2.1 Conceptual Issues**]
## .orange[**2.2 Simple vs. Complex Measures**]
## .orange[**2.3 Absolute and Relative Inequality**]
## .orange[**2.4 Weighting**]
## .orange[**2.5 Reference Points**]
]

---
class: center, top, inverse
# .orange[**2. Measuring Inequality**]

.left[
## .orange[**2.1 Conceptual Issues**]
## .gray[**2.2 Simple vs. Complex Measures**]
## .gray[**2.3 Absolute and Relative Inequality**]
## .gray[**2.4 Weighting**]
## .gray[**2.5 Reference Points**]
]

---
## Why monitor health inequalities?
.pull-left[
### Surveillance

- Natural complement to monitoring overall health

- Essential for detecting important changes in risk
]

--

.pull-right[
### Impact

- Opportunity to evaluate etiological explanations for health inequalities

- Evaluating the distributional impacts of public health interventions and medical innovations

- Crucial for measuring the responsiveness of health care systems to those most in need
]


---
class: center, top, inverse
# .orange[**Break!**] `r emo::ji("coffee")`
```{r, echo=F}
countdown(minutes = 15, 
          left = 0, right = 0, bottom = "15%", top = "15%",
          padding = "50px",
          margin = "5%",
          font_size = "7em",
          color_text= '#f5bc6c')
```