---
title: "The Gap-Closing Estimand"
subtitle: "Replicating Lundberg (2024) manually"
author: "Sam Harper"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    theme: cosmo
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## Overview

@lundberg2024 introduces the **gap-closing estimand**: a way to ask *how
much of an observed gap between social groups would close if we
intervened on a treatment?* This document works through the mechanics
using the package's own simulated dataset — first calling `gapclosing()`, then replicating each step 'by hand'.

The three manual estimators we cover are:

1.  **Outcome model** (g-computation / standardization)
2.  **Inverse probability weighting (IPW)**
3.  **Doubly-robust (AIPW)**

```{r}
#| label: setup
#| message: false
#| warning: false
library(here)
library(gapclosing)
library(tidyverse)
library(tinytable)
library(marginaleffects)
library(ggdag)
library(dagitty)
library(ggraph)
```

------------------------------------------------------------------------

## The causal model

Before touching data, we need to be explicit about the causal structure
we assume. The DAG below encodes our assumptions:

```{r}
#| label: dag
#| fig-width: 6
#| fig-height: 3
#| fig-align: center
#| echo: false
#| warning: false

dag <- dagify(
  T ~ L + X,
  Y ~ T + L + X,
  L ~ X,
  coords = list(
    x = c(X = 1, L = 2.5, T = 2.5, Y = 4),
    y = c(X = 1, L = 2,   T = 0,   Y = 1)
  ),
  labels = c(
    X = "Category\n(X)",
    L = "Confounder\n(L)",
    T = "Treatment\n(T)",
    Y = "Outcome\n(Y)"
  )
)

dag |>
  tidy_dagitty() |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(
    arrow = grid::arrow(length = grid::unit(6, "pt"), type = "closed"),
    edge_colour = "black", edge_width = 0.6
  ) +
  geom_dag_point(color = "white", size = 18, stroke = 1,
                 shape = 21, fill = "white") +
  geom_dag_label(aes(label = label), colour = "black", size = 3,
                 fontface = "plain", fill = NA, label.size = 0) +
  theme_dag_blank() +
  theme(plot.background = element_rect(fill = "white", color = NA))

```

**Reading the DAG:**

- **X → L**: Category membership determines the distribution of the
  confounder (in the simulation, L's mean shifts by −1.5 / 0 / +1.5
  across categories A / B / C).
- **L → T**: The confounder raises treatment probability (p =
  plogis(L)), so higher-L groups are more likely to be treated.
- **L → Y**: The confounder also directly raises the outcome (Y = L +
  ...), creating a backdoor path T ← L → Y.
- **T → Y**: The treatment effect of interest — and crucially, its
  magnitude *varies by category* (τ_A = +1.3, τ_B = 0, τ_C = −1.2).

The **backdoor path** T ← L → Y means a naive comparison of treated vs.
untreated confounds the treatment effect with baseline L differences.
Conditioning on L (within category) blocks this path and identifies
E\[Y(t)\].

------------------------------------------------------------------------

## The simulated data

The package ships a data-generating function. Here is exactly what it
does:

```{r}
#| label: dgp
set.seed(08544)
n <- 1000

dat <- data.frame(
  category = sample(c("A", "B", "C"), n, replace = TRUE)
) |>
  mutate(
    # L | X ~ N(mu_X, 1);  mu differs by category
    confounder = rnorm(
      n,
      mean = case_when(category == "A" ~ -1.5,
                       category == "B" ~  0,
                       category == "C" ~  1.5),
      sd = 1
    ),
    # T | L ~ Bernoulli(plogis(L))  — L confounds T -> Y
    treatment = rbinom(n, 1, plogis(confounder)),
    # category-specific treatment effects
    tau = case_when(category == "A" ~  1.3,
                    category == "B" ~  0.0,
                    category == "C" ~ -1.2),
    # Y | L, T, X ~ N(L + T*tau_X, 1)
    outcome = rnorm(n, 
      mean = confounder + treatment * tau, 
      sd = 1)
  ) |>
  select(-tau)   # not observed in practice

head(dat)
```

The key design features:

- **Confounding:** `confounder` (L) affects both treatment take-up
  (`plogis(L)`) and the outcome (`Y = L + ...`). Groups with higher L
  means (C \> B \> A) are both more likely to be treated *and* have
  higher baseline outcomes.
- **Effect heterogeneity:** τ is positive in A (+1.3), zero in B, and
  negative in C (−1.2). This means the *same* intervention (T=1) will
  *raise* outcomes in A but *lower* them in C.

------------------------------------------------------------------------

## Step 1 — The gap

Before any counterfactual reasoning, what group differences do we actually observe?

```{r}
#| label: factual
factual <- dat |>
  group_by(category) |>
  summarise(mean_y = mean(outcome), n = n(), .groups = "drop")

tt(factual, digits = 2)
```

```{r}
#| label: factual-gap
gap_AB_fact <- factual$mean_y[factual$category == "A"] -
               factual$mean_y[factual$category == "B"]
gap_CB_fact <- factual$mean_y[factual$category == "C"] -
               factual$mean_y[factual$category == "B"]

cat("Factual gap  A - B:", round(gap_AB_fact, 3),
    "\nFactual gap  C - B:", round(gap_CB_fact, 3), "\n")
```

These observed gaps conflate two things: (a) the effect of
treatment and (b) differences in the confounder L across categories. Group C has the highest L mean, so its members get treated more often *and* have higher baseline outcomes — even though τ_C is actually negative.

------------------------------------------------------------------------

## Step 2 — What are we actually estimating?

Before fitting any models it is worth being precise about the estimand,
because the gap-closing estimand is easy to confuse with a treatment
effect.

**What it is not.** A treatment effect contrast asks: for people in
group $x$, how much does treatment help?

$$E[Y(1) \mid X=x] - E[Y(0) \mid X=x] \quad \text{(ATE within category)}$$

**What it is.** The gap-closing estimand asks: after a common
intervention that sets $T=1$ for everyone, how much inequality *between*
groups remains?

$$E[Y(1) \mid X=A] - E[Y(1) \mid X=B] \quad \text{(gap-closing estimand)}$$

The comparison is *across groups under a shared counterfactual*, not
*across treatment arms within a group*. You hold the treatment fixed at
$T=1$ for everyone and ask how the outcome means still differ by
category.

The connection to the factual gap is what makes this useful for health
equity work. Start with the observed gap:

$$\underbrace{E[Y \mid X=A] - E[Y \mid X=B]}_{\text{factual gap}}$$

This conflates differential treatment rates, confounder differences, and
treatment effect heterogeneity. The gap-closing estimand isolates one
specific mechanism: *if we equalized treatment uptake at $T=1$, what
share of that gap would close?*

The answer need not be "most of it." In this simulation the gap nearly
closes because it is almost entirely driven by differential treatment
rates (through L). But equalizing treatment could leave most of the gap
intact — if the gap is driven by something treatment does not touch —
which is itself a substantive finding.

------------------------------------------------------------------------

## Step 3 — The `gapclosing()` function  {#sec-pkg}

**Counterfactual intervention:** set T = 1 for everyone.

The `gapclosing()` function fits an outcome model and a propensity model, combines them in a doubly-robust estimator, and returns the counterfactual mean for each category.

```{r}
#| label: gapclosing-pkg
#| message: false
#| warning: false
#| cache: true
estimate <- gapclosing(
  data                     = dat,
  counterfactual_assignments = 1,
  outcome_formula          = formula(outcome ~ confounder + category * treatment),
  treatment_formula        = formula(treatment ~ confounder + category),
  category_name            = "category",
  se                       = FALSE
)

summary(estimate)
```

Now let's recreate those numbers ourselves.


## Step 4 — By hand

The identification assumption throughout is **no unmeasured
confounding** given (category, confounder):

$$Y(t) \perp\!\!\!\perp T \mid \{\text{category},\, L\}$$

Because we generated the data, we know this holds by construction.

### 3a. Outcome model (standardization)

**Logic:** fit an outcome model on the observed data, then predict every person's outcome *as if* they had been treated (T = 1), and average within each category. This "standardizes" over the confounder distribution.

$$\hat{E}[Y(1) \mid X = x] = \frac{1}{n_x} \sum_{i: X_i = x} \hat{E}[Y \mid L_i, X_i, T = 1]$$

```{r}
#| label: om
# Fit outcome model
m_out <- lm(outcome ~ confounder + category * treatment, data = dat)

# Predict Y(T=1) for every person
dat_t1  <- dat |> mutate(treatment = 1)
dat$yhat_t1 <- predict(m_out, newdata = dat_t1)

# Average within category
cf_om <- dat |>
  group_by(category) |>
  summarise(cf_mean_om = mean(yhat_t1), .groups = "drop")

tt(cf_om, digits = 2)
```

This can also be accomplished with `marginaleffects`, but it takes some care in estimating. Using the `var` option with `datagrid=1` set the entire sample to a specific category then generates predictions and averages them:

```{r}
#| label: out_mp

cf_om_mp <- avg_predictions(m_out, 
  var = "category",
  datagrid(treatment=1)) |>
  select(category, estimate, conf.low, conf.high)

tt(cf_om_mp, digits = 2)
```

What we want, in fact, is specific predictions within each category while setting the treatment indicator to 1 for everyone, regardless of their actual treatment status: 

```{r}
#| label: out_mp2

tt(cf_om_mp, digits = 2)

cf_om_mp2 <- avg_predictions(m_out, 
  by = "category", 
  newdata = transform(dat, treatment = 1)) |>
  select(category, estimate, conf.low, conf.high)

tt(cf_om_mp2, digits = 2)

```


The outcome model is consistent when the outcome model is correctly specified. It can fail if the model is misspecified (wrong functional form, missing interactions).

### 3b. Inverse probability weighting (IPW)

**Logic:** reweight treated observations so that, within each category, the treated look like a random draw from the full population. Treated people with a low propensity (lucky to be treated) get upweighted; those with a high propensity get downweighted. Once again, there is more than one way to accomplish this.

One way is to use the **Hajek (stabilized) estimator** to keep weights well-behaved:

$$\hat{E}[Y(1) \mid X = x] = \frac{\sum_{i: X_i = x} \frac{T_i Y_i}{\hat{p}_i}}{\sum_{i: X_i = x} \frac{T_i}{\hat{p}_i}}$$

where $\hat{p}_i = \hat{P}(T = 1 \mid L_i, X_i)$.

Then one can just estimate the weighted means by category (weighted by IPW):

```{r}
#| label: ipw
# Fit propensity model
m_ps  <- glm(treatment ~ confounder + category,
             data = dat, family = binomial)
dat$ps <- predict(m_ps, type = "response")

# Hajek IPW estimator within each category
cf_ipw <- dat |>
  group_by(category) |>
  summarise(
    cf_mean_ipw = sum(treatment * outcome / ps) /
                  sum(treatment / ps),
    .groups = "drop"
  )

cf_ipw
```

Alternatively, one can also pursue a regression-based approach using stabilized weights:

```{r}
# Marginal treatment probability (scalar)
p_marginal <- mean(dat$treatment)

# Stabilized weights for treated observations
dat <- dat |>
  mutate(sw = if_else(treatment == 1, 
    p_marginal / ps, NA_real_))

# Weighted regression on treated only
m_ipw_stab <- lm(outcome ~ category, 
                 data   = dat[dat$treatment == 1, ],
                 weights = sw)

coef(m_ipw_stab)
```

IPW is consistent when the propensity model is correctly specified. It
can be noisy when propensity scores are near 0 or 1 (extreme weights).

### 3c. Doubly-robust (AIPW)

**Logic:** combine both models so that you only need *one* of them to be
correct. The AIPW pseudo-outcome for person $i$ is:

$$\psi_i = \underbrace{\hat{E}[Y \mid L_i, X_i, T=1]}_{\text{outcome model prediction}} + \underbrace{\frac{T_i}{\hat{p}_i}\bigl(Y_i - \hat{E}[Y \mid L_i, X_i, T=1]\bigr)}_{\text{IPW residual correction}}$$

Then $\hat{E}[Y(1) \mid X = x] = \bar{\psi}$ within category $x$.

When the outcome model is correct, the second term has mean zero — it's
just noise that washes out. When the propensity model is correct, the
IPW term corrects any bias from outcome model misspecification.

```{r}
#| label: dr
dat <- dat |>
  mutate(
    psi = yhat_t1 + (treatment / ps) * (outcome - yhat_t1)
  )

cf_dr <- dat |>
  group_by(category) |>
  summarise(cf_mean_dr = mean(psi), .groups = "drop")

cf_dr
```

------------------------------------------------------------------------

## Step 5 — Compare all estimators

```{r}
#| label: compare
results <- factual |>
  rename(factual = mean_y) |>
  select(category, factual) |>
  left_join(cf_om,  by = "category") |>
  left_join(cf_ipw, by = "category") |>
  left_join(cf_dr,  by = "category")

results
```

### Gap-closing fractions

How much of the factual gap (relative to category B) would close under
universal treatment?

```{r}
#| label: gap-closing
gcf <- function(cf_col) {
  cf <- results[[cf_col]]
  names(cf) <- results$category
  gap_cf <- cf["A"] - cf["B"]
  gap_f  <- results$factual[results$category == "A"] -
            results$factual[results$category == "B"]
  1 - gap_cf / gap_f
}

tibble(
  Estimator       = c("Outcome model", "IPW", "Doubly-robust"),
  `Gap A-B closed` = map_dbl(c("cf_mean_om","cf_mean_ipw","cf_mean_dr"), gcf)
)
```

```{r}
#| label: plot
#| fig-width: 8
#| fig-height: 4
plot_df <- results |>
  pivot_longer(-category, names_to = "estimator", values_to = "mean") |>
  mutate(estimator = recode(estimator,
    factual     = "Factual",
    cf_mean_om  = "Outcome model",
    cf_mean_ipw = "IPW",
    cf_mean_dr  = "Doubly-robust"
  ) |> factor(levels = c("Factual","Outcome model","IPW","Doubly-robust")))

ggplot(plot_df, aes(x = category, y = mean,
                    color = estimator, group = estimator)) +
  geom_point(size = 3) +
  geom_line() +
  scale_color_manual(values = c("black","#E41A1C","#377EB8","#4DAF4A")) +
  labs(
    x = "Category", y = "Mean outcome", color = NULL,
    title = "Factual vs. counterfactual (T = 1) outcome means",
    subtitle = "Three manual estimators vs. the gapclosing() package"
  ) +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

## Key takeaways

1.  **The factual gap is confounded.** Category C has the highest
    observed mean despite having a *negative* treatment effect (τ_C =
    −1.2), because C members have the highest baseline L.

2.  **Universal treatment (T = 1) nearly equalizes outcomes** across
    categories. The outcome model, IPW, and doubly-robust estimators all
    agree: the counterfactual means cluster much closer together than
    the factual means.

3.  **All three estimators agree here** because both the outcome model
    and the propensity model are correctly specified (we generated the
    data). In practice, the doubly-robust estimator is preferred because
    it only needs one of the two models to be right.

4.  **The gap-closing fraction** tells you what share of the observed
    gap a given intervention would eliminate — a more policy-relevant
    quantity than the raw treatment effect.

------------------------------------------------------------------------

## References

::: {#refs}
:::
