---
title: "The Gap-Closing Estimand: From Black Box to By Hand"
subtitle: "Replicating Lundberg (2021) manually"
author: "Sam Harper"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    theme: cosmo
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## Overview

@lundberg2024 introduces the **gap-closing estimand**: a way to ask *how
much of an observed gap between social groups would close if we
intervened on a treatment?* This document works through the mechanics
using the package's own simulated dataset — first calling `gapclosing()`
as a black box, then replicating each step by hand.

The three manual estimators we cover are:

1.  **Outcome model** (g-computation / standardization)
2.  **Inverse probability weighting (IPW)**
3.  **Doubly-robust (AIPW)**

```{r}
#| label: setup
#| message: false
#| warning: false
library(gapclosing)
library(tidyverse)
library(knitr)
```

------------------------------------------------------------------------

## The causal model

Before touching data, we need to be explicit about the causal structure
we assume. The DAG below encodes our assumptions:

```{r}
#| label: dag
#| fig-width: 5
#| fig-height: 1.8
#| fig-align: center
#| echo: false
#| warning: false

# Pure ggplot2 — same approach as Lundberg's own vignette.
nodes <- data.frame(
  label = c("Category\n(X)", "Confounder\n(L)", 
    "Treatment\n(T)", "Outcome\n(Y)"),
  x     = c(1, 2, 3, 4),
  y     = c(1, 1, 1, 1)
)

arr <- arrow(length = unit(0.1, "in"))

ggplot(nodes, aes(x = x, y = y, label = label)) +
  # Category → Confounder (straight)
  annotate("segment", x = 1, y = 1, xend = 1.6, yend = 1, arrow = arr) +
  # Category → Treatment (curved over)
  annotate("curve", x = 1, y = 1, xend = 2.9, yend = 1.2,
           arrow = arr, curvature = -0.3) +
  # Category → Outcome (curved over)
  annotate("curve", x = 1, y = 1, xend = 3.9, yend = 1.2,
           arrow = arr, curvature = -0.25) +
  # Confounder → Treatment (straight)
  annotate("segment", x = 2, y = 1, xend = 2.65, yend = 1, arrow = arr) +
  # Confounder → Outcome (curved under — the backdoor path)
  annotate("curve", x = 2, y = 1, xend = 3.75, yend = 0.82,
           arrow = arr, curvature = 0.3) +
  # Treatment → Outcome (straight, highlighted in blue)
  annotate("segment", x = 3, y = 1, xend = 3.65, yend = 1,
           color = "blue", linewidth = 1.1, arrow = arr) +
  geom_label(fill = "white", label.size = 0.3) +
  ylim(0.5, 1.5) +
  xlim(0.5, 4.8) +
  theme_void()
```

**Reading the DAG:**

- **X → L**: Category membership determines the distribution of the
  confounder (in the simulation, L's mean shifts by −1.5 / 0 / +1.5
  across categories A / B / C).
- **L → T**: The confounder raises treatment probability (p =
  plogis(L)), so higher-L groups are more likely to be treated.
- **L → Y**: The confounder also directly raises the outcome (Y = L +
  ...), creating a backdoor path T ← L → Y.
- **T → Y**: The treatment effect of interest — and crucially, its
  magnitude *varies by category* (τ_A = +1.3, τ_B = 0, τ_C = −1.2).

The **backdoor path** T ← L → Y means a naive comparison of treated vs.
untreated confounds the treatment effect with baseline L differences.
Conditioning on L (within category) blocks this path and identifies
E\[Y(t)\].

------------------------------------------------------------------------

## The simulated data

The package ships a data-generating function. Here is exactly what it
does:

```{r}
#| label: dgp
set.seed(08544)
n <- 1000

dat <- data.frame(
  category = sample(c("A", "B", "C"), n, replace = TRUE)
) |>
  mutate(
    # L | X ~ N(mu_X, 1);  mu differs by category
    confounder = rnorm(
      n,
      mean = case_when(category == "A" ~ -1.5,
                       category == "B" ~  0,
                       category == "C" ~  1.5),
      sd = 1
    ),
    # T | L ~ Bernoulli(plogis(L))  — L confounds T -> Y
    treatment = rbinom(n, 1, plogis(confounder)),
    # category-specific treatment effects
    tau = case_when(category == "A" ~  1.3,
                    category == "B" ~  0.0,
                    category == "C" ~ -1.2),
    # Y | L, T, X ~ N(L + T*tau_X, 1)
    outcome = rnorm(n, mean = confounder + treatment * tau, sd = 1)
  ) |>
  select(-tau)   # not observed in practice

head(dat)
```

The key design features:

- **Confounding:** `confounder` (L) affects both treatment take-up
  (`plogis(L)`) and the outcome (`Y = L + ...`). Groups with higher L
  means (C \> B \> A) are both more likely to be treated *and* have
  higher baseline outcomes.
- **Effect heterogeneity:** τ is positive in A (+1.3), zero in B, and
  negative in C (−1.2). This means the *same* intervention (T=1) will
  *raise* outcomes in A but *lower* them in C.

------------------------------------------------------------------------

## Step 1 — The factual gap

Before any counterfactual reasoning, what do we actually observe?

```{r}
#| label: factual
factual <- dat |>
  group_by(category) |>
  summarise(mean_y = mean(outcome), n = n(), .groups = "drop")

factual |> kable(digits = 3)
```

```{r}
#| label: factual-gap
gap_AB_fact <- factual$mean_y[factual$category == "A"] -
               factual$mean_y[factual$category == "B"]
gap_CB_fact <- factual$mean_y[factual$category == "C"] -
               factual$mean_y[factual$category == "B"]

cat("Factual gap  A - B:", round(gap_AB_fact, 3),
    "\nFactual gap  C - B:", round(gap_CB_fact, 3), "\n")
```

These observed gaps conflate two things: (a) the causal effect of
treatment and (b) differences in the confounder L across categories.
Group C has the highest L mean, so its members get treated more often
*and* have higher baseline outcomes — even though τ_C is actually
negative.

------------------------------------------------------------------------

## Step 2 — The `gapclosing()` black box

**Counterfactual intervention:** set T = 1 for everyone.

The `gapclosing()` function fits an outcome model and a propensity
model, combines them in a doubly-robust estimator, and returns the
counterfactual mean for each category.

```{r}
#| label: gapclosing-pkg
#| message: false
estimate <- gapclosing(
  data                     = dat,
  counterfactual_assignments = 1,
  outcome_formula          = formula(outcome ~ confounder + category * treatment),
  treatment_formula        = formula(treatment ~ confounder + category),
  category_name            = "category",
  se                       = FALSE   # skip bootstrap for speed
)

summary(estimate)
```

Now let's recreate those numbers ourselves.

------------------------------------------------------------------------

## Step 3 — By hand

The identification assumption throughout is **no unmeasured
confounding** given (category, confounder):

$$Y(t) \perp\!\!\!\perp T \mid \{\text{category},\, L\}$$

Because we generated the data, we know this holds by construction.

### 3a. Outcome model (g-computation)

**Logic:** fit an outcome model on the observed data, then predict every
person's outcome *as if* they had been treated (T = 1), and average
within each category. This "standardizes" over the confounder
distribution.

$$\hat{E}[Y(1) \mid X = x] = \frac{1}{n_x} \sum_{i: X_i = x} \hat{E}[Y \mid L_i, X_i, T = 1]$$

```{r}
#| label: om
# Fit outcome model
m_out <- lm(outcome ~ confounder + category * treatment, data = dat)

# Predict Y(T=1) for every person
dat_t1  <- dat |> mutate(treatment = 1)
dat$yhat_t1 <- predict(m_out, newdata = dat_t1)

# Average within category
cf_om <- dat |>
  group_by(category) |>
  summarise(cf_mean_om = mean(yhat_t1), .groups = "drop")

cf_om |> kable(digits = 3)
```

The outcome model is consistent when the outcome model is correctly
specified. It can fail if the model is misspecified (wrong functional
form, missing interactions).

### 3b. Inverse probability weighting (IPW)

**Logic:** reweight treated observations so that, within each category,
the treated look like a random draw from the full population. Treated
people with a low propensity (lucky to be treated) get upweighted; those
with a high propensity get downweighted.

We use the **Hajek (stabilized) estimator** to keep weights
well-behaved:

$$\hat{E}[Y(1) \mid X = x] = \frac{\sum_{i: X_i = x} \frac{T_i Y_i}{\hat{p}_i}}{\sum_{i: X_i = x} \frac{T_i}{\hat{p}_i}}$$

where $\hat{p}_i = \hat{P}(T = 1 \mid L_i, X_i)$.

```{r}
#| label: ipw
# Fit propensity model
m_ps  <- glm(treatment ~ confounder + category,
             data = dat, family = binomial)
dat$ps <- predict(m_ps, type = "response")

# Hajek IPW estimator within each category
cf_ipw <- dat |>
  group_by(category) |>
  summarise(
    cf_mean_ipw = sum(treatment * outcome / ps) /
                  sum(treatment / ps),
    .groups = "drop"
  )

cf_ipw |> kable(digits = 3)
```

IPW is consistent when the propensity model is correctly specified. It
can be noisy when propensity scores are near 0 or 1 (extreme weights).

### 3c. Doubly-robust (AIPW)

**Logic:** combine both models so that you only need *one* of them to be
correct. The AIPW pseudo-outcome for person $i$ is:

$$\psi_i = \underbrace{\hat{E}[Y \mid L_i, X_i, T=1]}_{\text{outcome model prediction}} + \underbrace{\frac{T_i}{\hat{p}_i}\bigl(Y_i - \hat{E}[Y \mid L_i, X_i, T=1]\bigr)}_{\text{IPW residual correction}}$$

Then $\hat{E}[Y(1) \mid X = x] = \bar{\psi}$ within category $x$.

When the outcome model is correct, the second term has mean zero — it's
just noise that washes out. When the propensity model is correct, the
IPW term corrects any bias from outcome model misspecification.

```{r}
#| label: dr
dat <- dat |>
  mutate(
    psi = yhat_t1 + (treatment / ps) * (outcome - yhat_t1)
  )

cf_dr <- dat |>
  group_by(category) |>
  summarise(cf_mean_dr = mean(psi), .groups = "drop")

cf_dr |> kable(digits = 3)
```

------------------------------------------------------------------------

## Step 4 — Compare all estimators

```{r}
#| label: compare
results <- factual |>
  rename(factual = mean_y) |>
  select(category, factual) |>
  left_join(cf_om,  by = "category") |>
  left_join(cf_ipw, by = "category") |>
  left_join(cf_dr,  by = "category")

results |> kable(digits = 3,
  col.names = c("Category", "Factual", "Outcome model", "IPW", "Doubly-robust"))
```

### Gap-closing fractions

How much of the factual gap (relative to category B) would close under
universal treatment?

```{r}
#| label: gap-closing
gcf <- function(cf_col) {
  cf <- results[[cf_col]]
  names(cf) <- results$category
  gap_cf <- cf["A"] - cf["B"]
  gap_f  <- results$factual[results$category == "A"] -
            results$factual[results$category == "B"]
  1 - gap_cf / gap_f
}

tibble(
  Estimator       = c("Outcome model", "IPW", "Doubly-robust"),
  `Gap A-B closed` = map_dbl(c("cf_mean_om","cf_mean_ipw","cf_mean_dr"), gcf)
) |>
  kable(digits = 3)
```

```{r}
#| label: plot
#| fig-width: 8
#| fig-height: 4
plot_df <- results |>
  pivot_longer(-category, names_to = "estimator", values_to = "mean") |>
  mutate(estimator = recode(estimator,
    factual     = "Factual",
    cf_mean_om  = "Outcome model",
    cf_mean_ipw = "IPW",
    cf_mean_dr  = "Doubly-robust"
  ) |> factor(levels = c("Factual","Outcome model","IPW","Doubly-robust")))

ggplot(plot_df, aes(x = category, y = mean,
                    color = estimator, group = estimator)) +
  geom_point(size = 3) +
  geom_line() +
  scale_color_manual(values = c("black","#E41A1C","#377EB8","#4DAF4A")) +
  labs(
    x = "Category", y = "Mean outcome", color = NULL,
    title = "Factual vs. counterfactual (T = 1) outcome means",
    subtitle = "Three manual estimators vs. the gapclosing() package"
  ) +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

## Key takeaways

1.  **The factual gap is confounded.** Category C has the highest
    observed mean despite having a *negative* treatment effect (τ_C =
    −1.2), because C members have the highest baseline L.

2.  **Universal treatment (T = 1) nearly equalizes outcomes** across
    categories. The outcome model, IPW, and doubly-robust estimators all
    agree: the counterfactual means cluster much closer together than
    the factual means.

3.  **All three estimators agree here** because both the outcome model
    and the propensity model are correctly specified (we generated the
    data). In practice, the doubly-robust estimator is preferred because
    it only needs one of the two models to be right.

4.  **The gap-closing fraction** tells you what share of the observed
    gap a given intervention would eliminate — a more policy-relevant
    quantity than the raw treatment effect.

------------------------------------------------------------------------

## References

::: {#refs}
:::
